---
title: "BIOL 416 R Tutorial 5 - Biodiversity-ecosystem function in plant communities"
author: "Mairin Deith"
output:
  html_document:
    highlight: zenburn
    theme: spacelab
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---

```{r, include=F}
setwd("~/Documents/Teaching/416_Spring19/Tutorials/BEFTutorial/Phylo/")
```

Read in packages and input files:
```{r}
library(ape)
tree<-read.tree("phylogeny.phy")

bef.data<-read.csv("../BEF2.csv", header=T)
```

### Setup the for-loop and dataframe to store results

```{r}
plots<-unique(bef.data$Plot)

#!# AAB = average above-ground biomass
results.df <- data.frame(
  Plots = plots,
  SR = NaN,
  PD = NaN,
  PD.trans = NaN,
  AAB = NaN
)
```

Practice for plot X2:
```{r}
sub.data <- subset(bef.data, Plot == "X2")
drop.spp <- colnames(sub.data)[
  which(sub.data[1, ]==0)
]
keep.spp <- colnames(sub.data)[
  which(sub.data[1, ]==1)
]
```

Use `drop.tip(...)` function to drop the species with 0 recorded in the BEF2.csv:
```{r}
phylo2 <- drop.tip(phy = tree, tip = drop.spp)
plot(phylo2)
```

**Note: extra species are included in the phylogeny along with the OUTGROUP!** 
These species are included in the phylogeny but were not present in the BEF2.csv (perhaps explaining why there are communities with zero species but some AAB).

Compare these species to:
```{r}
keep.spp
```

### For-loop 

```{r}
time.tree <- chronoMPL(tree)

for (i in 1:length(plots)){
  sub.data<-subset(bef.data,Plot==plots[i])
  species<-sub.data[,3:20]
  spp<-colnames(species)[species[1,]==1]
  results.df$SR[i]<-length(spp)
  results.df$AAB[i]<-mean(sub.data$AbvBioAnnProd)

# prune species NOT in the plots form the phylogeny
  subtree.s<-drop.tip(tree, tree$tip.label[!tree$tip.label %in% c(spp,"OUTGROUP")])
  subtree.t<-drop.tip(time.tree, time.tree$tip.label[!time.tree$tip.label %in% c(spp,"OUTGROUP")])
# note hear that we always include the OUTGROUP in the tree even though it is not included in any of the plots
# this is equivalent to adding a constant to the plot PD.
# It is necessary to do this, otherwise plots with species richness of one
# will crash the calculations of PD because you cannot have a tree with just one taxon (it is a stick!)

#sum edge lengths of pruned tree - this returns the phylogenetic diversity of included species
  results.df$PD[i]<-sum(subtree.s$edge.length)
  results.df$PD.trans[i]<-sum(subtree.t$edge.length)
}
```

### Models
**Jonathan removed plots with zero species (these still have biomass recorded because not all species were included in this dataframe)** - this is why there was a glut of values at the low end of PD. 
The plots below do not have the zeroes removed.

```{r}

# model.data<-subset(results.df, SR>0)
model.data <- results.df

model.PD.s<-lm(AAB~PD, data=as.data.frame(model.data))
model.PD.t<-lm(AAB~PD.trans, data=as.data.frame(model.data))
summary(model.PD.s)
```

### Plots
```{r}
plot(model.data$PD, model.data$AAB, xlab = "Phylogenetic Diversity (substitutions)", ylab = expression(paste("Biomass (g/m"^"2", ")")), cex = 1.25,cex.lab = 1.5, cex.axis = 1.25)
lines(model.data[,"PD"], predict(model.PD.s), lwd = 1.5)
```

```{r}
plot(model.data$PD.trans, model.data$AAB, xlab = "Phylogenetic Diversity (time)", ylab = expression(paste("Biomass (g/m"^"2", ")")), cex = 1.25,cex.lab = 1.5, cex.axis = 1.25)
lines(model.data[,"PD.trans"], predict(model.PD.t), lwd = 1.5)
```